Exploring the tree of numerical semigroups

### 2015-04-10 Papier/article.tex
Revisions asked by referees

\documentclass[reqno]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{pgfplots}
\usepackage{alltt}
\usepackage{listings}
\usepackage{xspace}
\usepackage{pdfsync}

@@

\title{Exploring the tree of numerical semigroups}
\author{Jean Fromentin and  Florent Hivert}
\date{3 November 2014}
\keywords{semigroups, tree, algorithm, \texttt{SSE}, 
\texttt{Cilk}, optimization}
\subjclass[2010]{05A15,68R05,68W10}

@@

\begin{abstract}
  In this paper we describe an algorithm visiting all numerical semigroups
  up to a given genus using a new representation. The
  interest of this algorithm is that it fits particularly well the
  architecture of modern computers allowing very large optimizations: we obtain the number of numerical semigroups of genus $g\leq  67$ and we confirm the Wilf conjecture for $g\leq 60$.

@@

For a given positive integer $g$, the number of numerical semigroups of genus $g$ is finite and is denoted by $n_g$. 
In  J.A. Sloane's \emph{on-line encyclopedia of integer sequences}~\cite{OEIS} we find the values of $n_g$ for $g\leq 52$. 
These values have been obtained by M. Bras-Amor\'os (\cite{BrasAmoros2008} for more details for $g\leq 50)$. 
On his home page~\cite{Delgado}, M. Delgado  gives the value of $n_{55}$. 
The values of $n_{53}$ and $n_{54}$ seem to be so far available nowhere.

@@

Here we describe a new algorithm for the exploration of the tree of numerical
semigroups $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a new combinatorial representation of
numerical semigroups that is well suited and allows large code optimization essentially based on the use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our new encoding of numerical semigroups and the associated
algorithms, and second to present the optimization techniques which allow, for
those kinds of algorithms, to get speedups by factors of hundreds and even
thousands. We claim that these techniques are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores.

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
The description of our new representation of numerical semigroups is done in the second section.
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimization of the algorithm introduced in Section~\ref{S:Algo}.
In the last section we emphasize the results obtained using our algorithm.

@@

We start this section with definitions and properties of numerical semigroups that will be used in the sequel.
For a more complete introduction, the reader can usefully consult the book \emph{Numerical Semigroups} of J.C.~Rosales and P.A.~Garc\'ia-S\'anchez \cite{BookNS} or the book \emph{The Diophantine Frobenius Problem} of J.L.~Ram\'irez~Alfons\'in \cite{BookDFP}.

@@

$iii)$ $c(S)=1+\max(\NN\setminus S)$, the \emph{conductor} of $S$ for $S$ different from $\NN$. By convention, the conductor of $\NN$ is $0$. 
\end{defi}

@@

\begin{lem}[Lemma 2.3, Chap. I, \cite{BookNS}]
For a numerical semigroup $S$, the set $\Irr(S)$ is the minimal generating set of $S$ relatively to the inclusion ordering.
\end{lem}

@@

\begin{prop}
\label{P:Res}
For every  numerical semigroup  $S$,  we have

@@

\begin{proof}
$i)$ Let $x$ be a positive integer satisfying $x \geq c(S)+m(S)$. 
As the integers $x$ and $x-m(S)$ are greater than $c(S)$, they belong to $S$.
Since $x$ equals $(x-m(S))+m(S)$ with $x-m(S)\not=0$ and $m(S)\not=0$, it cannot be irreducible.

$ii)$ The set $\NN\setminus{S}$ of cardinality $g(S)$ contains $\{1,..,m(S)-1\}$ and so the relation $m(S)\leq g(S)+1$ holds.

$iii)$ Let $x$ be an element of $S\cap\{0,...,c(S)-1\}$. 
The integer $y=c(S)-1-x$ lies in $\{0,...,c(S)-1\}$. 
Moreover $y$ is not in $S$, for otherwise  we write $c(S)-1=y+x$ where $x,y$ are elements of $S$ implying $c(S)-1$ belongs to $S$. 
In contradiction with the definition of conductor. 
Thus we define an involution 
\[
\begin{array}{rcl}
\psi:\{0,...,c(S)-1\}&\to&\{0,...,c(S)-1\}\\
x&\mapsto&c(S)-1-x
\end{array}
\]
that sends $S\{0,...,c(S)-1\}$ into $\{0,...,c(S)-1\}\setminus S$ which is  of cardinality~$g(S)$. 
Let us denote by $k$ the cardinality  of $S\cap\{0,...,c(S)-1\}$. 
Since $\psi$ is injective we must have $k\leq g(S)$. 
From the relation 
\[
\{0,...,c(S)-1\}=S\cap\{0,...,c(S)-1\}\sqcup\{0,...,c(S)-1\}\setminus S
\]
 we obtain $c(S)=k+g(S)$ and so $c(S)\leq 2 g(S)$.
\end{proof}

@@

We now explain the construction of the tree of numerical semigroups.
Let~$S$ be a numerical semigroup. The set $S'=S\cup\{c(S)-1\}$ is also a numerical  semigroup and its genus is $g(S)-1$. 
As  each integer greater than $c(S)-1$ is included in $S'$ we have $c(S')\leq c(S)-1$. 
Therefore every semigroup $S$ of genus $g$ can be obtained from a semigroup $S'$ of genus $g-1$ by removing an element of~$S'$ greater than or equal to $c(S')$.

\begin{prop}
\label{P:Sx}
Let $S$ be a numerical semigroup and $x$ an element of $S$. The set~$S^x=S\setminus\{x\}$ is a numerical semigroup if and only if $x$ is irreducible in $S$.
\end{prop}

\begin{proof}
If $x$ is not irreducible in  $S$, then there exist $a$ and $b$ in $S\setminus{0}$ such that $x=a+b$. 
As $a\not=0$ and $b\not=0$ hold, the integers  $a$ and $b$ belong to $S\setminus\{x\}$. 
Since $x=a+b$ and $x\not\in S^x$, it follows that  $S^x$ is not stable under addition.

Conversely, assume that $x$ is irreducible in  $S$. 
As $0$ is never irreducible, the set~$S^x$ contains~$0$. 
Let $a$ and $b$ be two integers belonging to $S^x$. 
The set~$S$ is stable under addition, hence $a+b$ lies in $S$. 
As $S$ is equal to $S^x\cup\{x\}$, the integer $a+b$ also lies in $S^x$ except if it is equal to $x$. 
The latter is impossible since $a$ and $b$ are different from $x$ and $x$ is irreducible.
\end{proof}

@@

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroups~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(s),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{g}$ the subtree of~$\mathcal{T}$ restricted to all semigroups of genus $\leq g$.

@@

As the reader can check, the main difficulty to characterize the sons of a semigroup is to determine its irreducible elements.
 In \cite{NumericalSgps}, the semigroups are given by their Ap\'ery sets and then the difficulty is to describe the Ap\'ery set of $S^x$  from that of $S$. 
 This approach is elegant but not sufficiently \emph{basic} for our optimizations.

@@

The aim of this section is to describe a new representation of numerical semigroups,
which is well suited to an efficient exploration of the tree~$\mathcal{T}$ of numerical semigroups.

@@

Assume that $y$ is an element of $D_S(x)$.
By very definition of $D_S(x)$, the integer~$z=x-y$ also belongs to $S$. 
Then $x$ can be decomposed as $x=y+z$ with $y$ and $z$ in~$S$. 
Moreover the condition $2y\leq x$ implies $y\leq z$. 
In other words if we define~$D'_S(x)$ to be the set of all $(y,z)\in S\times S$ with $x=y+z$ and $y\leq z$ then~$D_S(x)$ is the image of $D'_S(x)$ under the projection on the first coordinate.
 Hence~$D_S(x)$ describes how~$x$ can be decomposed as sums of two elements of $S$.
 This justifies the name given to the function $d_S$.

@@

@@

\begin{proof}
We start with $i)$. If $x$ is an element of $S$ then $x$ equals $0+x$. 
The relation~$2\times 0\leq x$ and $0\in S$ imply that $D_S(x)$ contains $0$, and so $d_S(x)>0$ holds.
Conversely, the relation  $d_S(x)>0$  implies that~$D_S(x)$ is non-empty. 
Let $y$ be an element of~$D_S(x)$. 
As $y$ and $x-y$ belong to  $S$, by definition, the integer $x=(x-y)+y$ is in $S$  (since $S$ is stable by addition).

Let us show $ii)$. 
Assume $x$ is irreducible in $S$.
There cannot exist $y$ and $z$ in~$(S\setminus\{0\})^2$ such that~$x=y+z$. 
The only possible decomposition of $x$ as a sum of two elements of $S$ is $x=0+x$. 
Hence, the set $D_S(x)$ is equal to~$\{0\}$ and we have $d_S(x)=1$. 
Conversely, let $x$ be such that $d_S(x)$ is $1$. 
By~$i)$ the  integer $x$ must be in $S$. 
As $x=0+x$ is always a decomposition of $x$ as a sum of two elements in $S$, we obtain $D_S(x)=\{0\}$.
If there exist $y$ and $z$ in $S$ such that $y\leq z$ and $x=y+z$ hold then $y$ lies in~$D_S(x)$. 
This implies $y=0$ and $z=x$. 
Hence $x$ is irreducible in~$S$.
\end{proof}

We note  that $0$ is never irreducible despite the fact $d_S(0)$ is $1$ for all numerical semigroups $S$.

We now explain how to compute the $S$-decomposition numbers function of a numerical semigroup  from that of its father.

@@

\begin{proof}
Let $y$ be in $\NN\setminus\{0\}$. We have 
\[
D_{S^x}(y)=\{z \in S^x\ |\ y-z\in S^x\ \text{and}\  2z\leq y\}.
\]
We construct two sets as follow :
\begin{align*}
E&=\{z\in S^x\ |\ y-z=x\ \text{and}\ 2z\leq y\}\\
F&=\begin{cases}
\{x\}&\text{if $y-x\in S$ and $2x\leq y$},\\
\emptyset&\text{otherwise}.
\end{cases}
\end{align*}
We have the relation $D_{S}(y)=D_{S^x}(y)\sqcup E\sqcup F$ and so $d_S(y)$ is equal to $d_{S^x}(y)+\card(E)+\card(F)$
and $D_{S^x}(y)$ is a subset of $D_S(y)$. We start by determining $E$. 
The relation $y-z=x$ implies $z=y-x$ and $2z\leq y$ becomes $y\leq 2x$. Thus we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S^x$ and $y\leq 2x$}\\
\emptyset&\text{otherwise}.
\end{cases}
\] 
The relation $y-x\in S^x$ together with $y\leq2x$ is equivalent to $y-x\in S$ together with $y<2x$.
Therefore we obtain
\[
E=\begin{cases}
\{y-x\}&\text{if $y-x\in S$ and $y< 2x$}\\
\emptyset&\text{otherwise}.
\end{cases}
\] 
Assume $y-x$ lies in $S$.  If $y<2x$ holds, we have $E=\{y-x\}$ and $F=\emptyset$. Otherwise, we obtain $E=\emptyset$ and $F=\{x\}$. 
Then if $y-x\in S$ holds, we have $\card(E)+\card(F)=1$ and so $d_{S}(y)$ is equal to $d_{S^x}(y)+1$.
If $y-x$ does not belong to $S$, then $E$ and $F$ are empty implying $d_{S}(y)=d_{S^x}(y)$. As the condition $y-x\in S$ is equivalent to $y\geq x$ together with $d_S(y-x)>0$, by Proposition~\ref{P:Dandd} the result holds.
\end{proof}

\section{A new algorithm}
\label{S:Algo}
We can easily explore the tree of numerical semigroups up to a genus~$G$ using
a depth first search algorithm (see Algorithm~\ref{A:ExpRec}). This approach does not seem to
have been used before.  In particular, M.~Bras-Amor\'os and M.~Delgado use
instead a breadth first search exploration.  The main advantage in our
approach is the small memory needs.  The cost to pay is that, if we want to
explore the tree deeper, we must restart from the root.

@@

\begin{prop}
\label{P:delta}
Let $G$ be an integer and $S$ be a numerical semigroup of genus $g\leq G$.
Then $S$ is entirely described by the vector $\delta_S=(d_S(0),...,d_S(3\,G)) \in \NN^{3G+1}$. More precisely we can obtain $c(S), g(S), m(S)$ and $\Irr(S)$ from $\delta_S$.
\end{prop}

\begin{proof}
By Proposition~\ref{P:Res} $iii)$ we have  the relation $c(S)\leq 2g(S)$ and so the $S$-decomposition number of~$c(S)$ occurs in $\delta_S$.
Since $c(S)$ is equal to $1+\max(\NN\setminus S)$ for $S\not=\NN$, Proposition~\ref{P:delta} implies 
\[
c(S)=1+\max\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
As all elements of $\NN\setminus S$ are smaller than $c(S)$, their $S$-decomposition numbers are in $\delta_S$ and we obtain
\[
g(S)=\card\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
By Proposition~\ref{P:Res} $ii)$, the relation $m(S)\leq g(S)+1$ holds. 
This implies that the $S$-decomposition number of $m(S)$ appears in $\delta_S$ : 
\[
m(S)=\min\{i\in\{0,...,3G\},\ d_S(i)>0\}.
\]
By Proposition~\ref{P:Res},  all irreducible elements are smaller than $c(S)+m(S)-1$, which is itself smaller than~$3G$.
Hence, Proposition~\ref{P:Dandd} gives
\[
\Irr(S)=\{i \in\{0,...,3G\},\ d_S(i)=1\}.
\]
\end{proof}

@@

\begin{algorithm}
\caption{\small Returns the son $S^x$ of $S$ with $x\in\Irr(S)\cap[c(S),c(S)+m(S)[$.}\label{A:Son}
\begin{algorithmic}[1]
\small
\Function{Son}{\tt{S},\tt{x},\tt{G}}
\State \tt{$\tt{S}^\tt{x}$.c} $\gets$ $\tt{x}+1$
\State \tt{$\tt{S}^\tt{x}$.g} $\gets$ $\tt{S.g}+1$
\If{$\tt{x}>\tt{S.m}$}
	\State \tt{$\tt{S}^\tt{x}$.m} $\gets$ \tt{S.m}
\Else
	\State \tt{$\tt{S}^\tt{x}$.m} $\gets$ $\tt{S.m}+1$
\EndIf
\State\tt{$\tt{S}^\tt{x}$.d} $\gets$ \tt{S.d}
\For{\tt{y} from \tt{x} to $3\,\tt{G}$}
	\If{$\tt{S.d}[\tt{y}-\tt{x}]>0$}
		\State $\tt{$\tt{S}^\tt{x}$.d}[\tt{y}]\gets\tt{$\tt{S}^\tt{x}$.d}[\tt{y}]-1$
	\EndIf
\EndFor
\State{\textbf{return} $\tt{S}^\tt{x}$}
\EndFunction
\end{algorithmic}
\end{algorithm}

\begin{prop}
\label{P:Son}
Running on $(S,x,G)$ with $g(S)\leq G$, $x\in \Irr(S)$ and~$x\leq c(S)$, Algorithm~\ref{A:Son} returns the semigroup $S^x$ in time $O(\log(G)\times G)$.
\end{prop}

\begin{proof}
Let us check the correctness of the algorithm.
By construction $S^x$ is the semigroup~$S\setminus\{x\}$.
Thus the genus of $S^x$ is $g(S)+1$, see Line~\tt{3}.
Every integer of $I=[x+1,+\infty[$ lies in $S$ since $x$ is greater than~$c(S)$, so the interval $I$ is included in $S^x$.
As $x$ does not belong to $S$, the conductor of $S^x$ is $x+1$, see Line~\tt{2}.
For the multiplicity of $S^x$ we have two cases. 
First, if $x>m(S)$ holds  then $m(S)$ is also in $S^x$ and so $m(S^x)$ is equal to $m(S)$.
Assume now $x=m(S)$. 
The relation $x(S)\geq c(S)$ and the characterization of~$m(S)$ implies $x=m(S)=c(S)$.
Thus $S^x$ contains $m(S)+1$ which is $m(S^x)$.
The initialization of~$m(S^x)$ is done by Lines \tt{4} to \tt{8}. 
The correctness of the computation of $\delta_{S^x}$ (see Proposition~\ref{P:delta} for the definition of $\delta_{S^x}$) done from Line \tt{9} to Line \tt{15} is a direct consequence of~Proposition~\ref{P:Dandd}.

@@

Let us now prove the space complexity statement.
For  this we need to describe the stack through  the run of the algorithm.
Since the stack is filled  with a depth first search algorithm, it has two properties.
The first one is that reading the stack from the bottom to the top, the genus increases.  
The second one is that, for all $g\in[0,G]$, every semigroup of genus~$g$ in the stack has the same father.
As the number of sons of a  semigroup $S$ is the number of $S$-irreducible elements in the set $\{c(S),...,c(S)+m(S)-1\}$,  a semigroup $S$ has at most $m(S)$ sons. 
By~Proposition~\ref{P:Res} $ii)$, this implies that a semigroup of genus $g$ as at most $g+1$ sons.
Therefore the stack contains at most $g+1$ semigroups of genus $g+1$ for $g\leq G$. 
So the size of the stack is bounded~by 
\[
M=\sum_{g=0}^Gg=\frac{G(G+1)}2.
\]
A semigroup $S$ is represented by a quadruple $(c(S),g(S),m(S),\delta_S)$. 
By relations $ii)$ and $iii)$ of Proposition~\ref{P:Res}, we have $c\leq 2g(S)$ and $m\leq g(S)+1$. 
As $g(S)\leq G$ holds, the integers~$c$, $g$  and $m$ of the representation of $S$ require a memory space in $O(\log(G))$. 
The size of~$\delta_S=(d_S(0),...,d_S(3G))$ is exactly~$3G+1$. 
Each entry of $\delta_S$ is the $S$-decomposition number of an integer smaller than $3G$ and hence
requires $O(\log(G))$ bytes of memory space.
Therefore the space complexity of $\delta_S$ is in~$O(\log(G)\times G)$, which implies that the space complexity of the \textsc{Count} algorithm is
\[
O(\log(G)\times G\times M)= O(\log(G)\times G^3).
\]
\end{proof}

@@

\item Vectorization (\MMX, \SSE instructions sets) and careful memory alignment. 
\item Shared memory multi-core computing using \CilkP for low level
  enumerating tree branching;
\item Partially derecursived algorithm using a stack;
\item Avoiding all dynamic allocation during the computation: everything is
  computed ``in place'';
\item Avoiding all unnecessary copy (the cost of the \textsc{Son} algorithm is
  roughly the same as copying) !
\item Aggressive loop unrolling: the main loop is unrolled by hand using some
  kind of Duff's device.
\item Careful choice of data type (\verb|uint_fast8_t| for decomposition
  number, vs \verb|uint_fast64_t| for all indexes).
\end{itemize}

The source code of our algorithm is available here \cite{code}.
\subsection{Vectorization}

@@

{\small
\begin{alltt}
start = x >> 4;      // index of the block containing x
shift = x & 0xF;     // offset of x inside the block
...                  // some specific instructions to
...                  // handle the first incomplete block.
for (long int i=start+1; i<NBLOCKS; i++) {
  block = load_unaligned_epi8(src + ((i-start)<<4) - shift);
  dst[i] -= ((block != zero) & one);
}
\end{alltt}
}

>>>>

\synctex=1
\documentclass[reqno]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{pgfplots}
\usepackage{alltt}
\usepackage{listings}
\usepackage{xspace}
%\usepackage{pdfsync}

@@

\title{Exploring the tree of numerical semigroups}
\author{Jean Fromentin and  Florent Hivert}
\date{April 10,  2014}
\keywords{semigroups, tree, algorithm, \texttt{SSE}, 
\texttt{Cilk}, optimization}
\subjclass[2010]{05A15,68R05,68W10}

@@

\begin{abstract}
  In this paper we describe an algorithm visiting all numerical semigroups
  up to a given genus using a well suited representation. The
  interest of this algorithm is that it fits particularly well the
  architecture of modern computers allowing very large optimizations: we obtain the number of numerical semigroups of genus $g\leq  67$ and we confirm the Wilf conjecture for $g\leq 60$.

@@

For a given positive integer $g$, the number of numerical semigroups of genus $g$ is finite and is denoted by $n_g$. 
In  J.A. Sloane's \emph{on-line encyclopedia of integer sequences}~\cite{OEIS} we find the values of $n_g$ for $g\leq 52$. 
These values have been obtained by M. Bras-Amor\'os (\cite{BrasAmoros2008} for more details for $g\leq 50)$. 
On his home page~\cite{Delgado}, M. Delgado  gives the value of $n_{55}$.

@@

Here we describe a new algorithm for the exploration of the tree of numerical
semigroups $\mathcal{T}$ and achieve the computation of $n_g$ for $g\leq 67$.
The cornerstone of our method is a combinatorial representation of
numerical semigroups that is well suited and allows large code optimization essentially based on the use of
vectorial instructions and parallelization.
The goal of the paper is twofold:
first to present our encoding of numerical semigroups and the associated
algorithms, and second to present the optimization techniques which allow, for
those kinds of algorithms, to get speedups by factors of hundreds and even
thousands. We claim that these techniques are fairly general for those kinds of
algorithms. As a support for the claim, we applied it to an algorithm of
N.~Borie enumerating integer vector modulo permutation groups~\cite{Borie} and
got a speedup by a factor larger than 2000 using 8 cores.

The paper is divided as follows.
In Section~\ref{S:Tree} we describe the tree of numerical semigroups and give bounds for some parameters attached to a numerical semigroup.
The description of our representation of numerical semigroups is done in the second section.
In Section~\ref{S:Algo} we describe an algorithm based on the representation given in Section~\ref{S:DecNumber} and give its complexity. 
Section~\ref{S:Opti} is more technical, and is devoted to the optimization of the algorithm introduced in Section~\ref{S:Algo}.
In the last section we emphasize the results obtained using our algorithm.

@@

We start this section with definitions and properties of numerical semigroups that will be used in the sequel.
For a more complete introduction, the reader can usefully consult the book \emph{Numerical Semigroups} by J.C.~Rosales and P.A.~Garc\'ia-S\'anchez \cite{BookNS} or the book \emph{The Diophantine Frobenius Problem} by J.L.~Ram\'irez~Alfons\'in \cite{BookDFP}.

@@

$iii)$ $f(S)=\max(\mathbb{Z}\setminus S)$, the \emph{Frobenius} of $S$;

$iv)$ $c(S)=f(S)+1$, the \emph{conductor of $S$}. 
\end{defi}

@@

\begin{lem}[Lemma 2.3 of~\cite{BookNS}]
For a numerical semigroup $S$, the set $\Irr(S)$ is the minimal generating set of $S$ relatively to the inclusion ordering.
\end{lem}

@@

\begin{prop}[Proposition~2.12 and Lemma~2.14 of~\cite{BookNS}]
\label{P:Res}
For every  numerical semigroup  $S$,  we have

@@

@@

We now explain the construction of the tree of numerical semigroups.
Let~$S$ be a numerical semigroup. The set $S'=S\cup\{f(S)\}$ is also a numerical  semigroup and its genus is $g(S)-1$. 
As  each integer greater than $f(S)$ is included in $S'$ we have $c(S')\leq f(S)$. 
Therefore every semigroup $S$ of genus $g$ can be obtained from a semigroup $S'$ of genus $g-1$ by removing an element of~$S'$ greater than or equal to $c(S')$.

\begin{prop}[Proposition~7.28 of \cite{BookNS}]
\label{P:Sx}
Let $S$ be a numerical semigroup and $x$ an element of $S$. The set~$S^x=S\setminus\{x\}$ is a numerical semigroup if and only if $x$ is irreducible in $S$.
\end{prop}

@@

We construct the tree of numerical semigroups, denoted by $\mathcal{T}$ as follows. 
The root of the tree is the unique semigroup of genus $0$, \ie, $\left<1\right>$ that is equal to $\NN$. 
If~$S$ is a semigroup in the tree,  the sons of $S$ are exactly the semigroups~$S^x$ where $x$ belongs to $\Irr(S)\cap[c(S),+\infty[$. 
By convention, when depicting the tree, the numerical semigroup $S^x$ is in the left of $S^y$ if $x$ is smaller than~$y$. 
With this construction, a semigroup $S$ has depth $g$ in $\mathcal{T}$ if and only if its genus is~$g$, see Figure~\ref{F:Tree}.
We denote by $\mathcal{T}_{g}$ the subtree of~$\mathcal{T}$ restricted to all semigroups of genus $\leq g$.

@@

@@

The aim of this section is to describe a representation of numerical semigroups,
which is well suited to an efficient exploration of the tree~$\mathcal{T}$ of numerical semigroups.

@@

Assume that $y$ is an element of $D_S(x)$.
By definition of $D_S(x)$, the integer~$z=x-y$ also belongs to $S$. 
Then $x$ can be decomposed as $x=y+z$ with $y$ and $z$ in~$S$. 
Moreover the condition $2y\leq x$ implies $y\leq z$. 
In other words if we define~$D'_S(x)$ to be the set of all $(y,z)\in S\times S$ with $x=y+z$ and $y\leq z$ then~$D_S(x)$ is the image of $D'_S(x)$ under the projection on the first coordinate.
 Hence~$D_S(x)$ describes how~$x$ can be decomposed as sums of two elements of $S$.
 This justifies the name given to the function $d_S$.

@@

A straightforward consequence of the definition of $S$-decomposition numbers is :

@@

We note  that $0$ is never irreducible despite the fact $d_S(0)$ is $1$ for all numerical semigroups $S$.
We now explain how to compute the $S$-decomposition numbers function of a numerical semigroup  from that of its father.

@@

\begin{proof}
A direct consequence of~$D_{S^x}(y)=D_S(y)\setminus\{y-x,x\}$.
\end{proof}

\section{A new algorithm}
\label{S:Algo}
We can easily explore the tree of numerical semigroups up to a genus~$G$ using
a depth first search algorithm (see Algorithm~\ref{A:ExpRec}). This approach does not seem to
have been used before.  In particular, M.~Bras-Amor\'os and M.~Delgado use
instead a breadth first search exploration.  The main advantage in our
approach is the small memory needs.  
%[MODIF]
Indeed, in the case of breadth first search algorithm one needs to compute and store the list $L_g$ of all numerical semigroups of genus $g$ before 
visiting the numerical semigroups of genus $g+1$, which is not required for a depth first search exploration.
In this paper we are interested in the exploration of the list~$L_g$ for $g\in \NN$, not in storing it.
This is the reason we use a depth first search algorithm for the exploration of the tree $\mathcal{T}_g$.
The only limitation is then the duration of the exploration and not the amount of available memory.
For example the list~$L_{54}$ needs several terabytes to be stored. 
%[/MODIF]

@@

\begin{prop}
\label{P:delta}
Let $G$ be an integer and $S$ be a numerical semigroup of genus $0<g\leq G$.
Then $S$ is entirely described by the vector $\delta_S=(d_S(0),...,d_S(3\,G)) \in \NN^{3G+1}$. More precisely we can obtain $c(S), g(S), m(S)$ and $\Irr(S)$ from $\delta_S$.
\end{prop}

\begin{proof}
By Proposition~\ref{P:Res} $iii)$ we have  the relation $c(S)\leq 2g(S)$ and so the $S$-decomposition number of~$c(S)$ occurs in $\delta_S$.
Proposition~\ref{P:delta} implies 
\[
c(S)=1+\max\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
As all elements of $\NN\setminus S$ are smaller than $c(S)$, their $S$-decomposition numbers are in $\delta_S$ and we obtain
\[
g(S)=\card\{i\in\{0,...,3G\},\ d_S(i)=0\}.
\]
By Proposition~\ref{P:Res} $ii)$, the relation $m(S)\leq g(S)+1$ holds. 
This implies that the $S$-decomposition number of $m(S)$ appears in $\delta_S$ : 
\[
m(S)=\min\{i\in\{0,...,3G\},\ d_S(i)>0\}.
\]
By Proposition~\ref{P:Res},  all irreducible elements are smaller than $c(S)+m(S)-1$, which is itself smaller than~$3G$.
Hence, Proposition~\ref{P:Dandd} gives
\[
\Irr(S)=\{i \in\{0,...,3G\},\ d_S(i)=1\}.\qedhere
\]
\end{proof}

The previous representation of numerical semigroup (in terms of the vector $\delta_S$)  is similar but a little different from this used in \cite{Maria} and by people concerned with coding theory.

@@

\begin{algorithm}
\caption{\small Returns the son $S^x$ of $S$ with $x\in\Irr(S)\cap[c(S),c(S)+m(S)[$.}\label{A:Son}
\begin{algorithmic}[1]
\small
\Function{Son}{\tt{S},\tt{x},\tt{G}}
\State \tt{$\tt{S}^\tt{x}$.c} $\gets$ $\tt{x}+1$
\State \tt{$\tt{S}^\tt{x}$.g} $\gets$ $\tt{S.g}+1$
\If{$\tt{x}>\tt{S.m}$}
	\State \tt{$\tt{S}^\tt{x}$.m} $\gets$ \tt{S.m}
\Else
	\State \tt{$\tt{S}^\tt{x}$.m} $\gets$ $\tt{S.m}+1$
\EndIf
\State\tt{$\tt{S}^\tt{x}$.d} $\gets$ \tt{S.d}\Comment{copy all the decomposition numbers}
\For{\tt{y} from \tt{x} to $3\,\tt{G}$}
	\If{$\tt{S.d}[\tt{y}-\tt{x}]>0$}
		\State $\tt{$\tt{S}^\tt{x}$.d}[\tt{y}]\gets\tt{$\tt{S}$.d}[\tt{y}]-1$ \Comment{decrease the decomposition number by $1$}
	\EndIf
\EndFor
\State{\textbf{return} $\tt{S}^\tt{x}$}
\EndFunction
\end{algorithmic}
\end{algorithm}

\begin{prop}
\label{P:Son}
Running on $(S,x,G)$ with $g(S)\leq G$, $x\in \Irr(S)$ and~$x\geq c(S)$, Algorithm~\ref{A:Son} returns the semigroup $S^x$ in time $O(\log(G)\times G)$.
\end{prop}

\begin{proof}
Let us check the correctness of the algorithm.
By construction $S^x$ is the semigroup~$S\setminus\{x\}$.
Thus the genus of $S^x$ is $g(S)+1$, see Line~\tt{3}.
Every integer of $I=[x+1,+\infty[$ lies in $S$ since $x$ is greater than~$c(S)$, so the interval $I$ is included in $S^x$.
As $x$ does not belong to $S^x$, the conductor of $S^x$ is $x+1$, see Line~\tt{2}.
For the multiplicity of $S^x$ we have two cases. 
First, if $x>m(S)$ holds  then $m(S)$ is also in $S^x$ and so $m(S^x)$ is equal to $m(S)$.
Assume now $x=m(S)$. 
The relation $x(S)\geq c(S)$ and the characterization of~$m(S)$ implies $x=m(S)=c(S)$.
Thus $S^x$ contains $m(S)+1$ which is $m(S^x)$.
The initialization of~$m(S^x)$ is done by Lines \tt{4} to \tt{8}. 
The correctness of the computation of $\delta_{S^x}$ (see Proposition~\ref{P:delta} for the definition of $\delta_{S^x}$) done from Line \tt{9} to Line \tt{15} is a direct consequence of~Proposition~\ref{P:Dandd}.

@@

Let us now prove the space complexity statement.
For  this we need to describe the stack through  the run of the algorithm.
Since the stack is filled  with a depth first search algorithm, it has two properties.
The first one is that reading the stack from the bottom to the top, the genus increases.  
The second one is that, for all $g\in[0,G]$, every semigroup of genus~$g$ in the stack has the same father.
As the number of sons of a  semigroup $S$ is the number of $S$-irreducible elements in the set $\{c(S),...,c(S)+m(S)-1\}$,  a semigroup $S$ has at most $m(S)$ sons. 
By~Proposition~\ref{P:Res} $ii)$, this implies that a semigroup of genus $g$ has at most $g+1$ sons.
Therefore the stack contains at most $g+1$ semigroups of genus $g+1$ for $g\leq G$. 
So the size of the stack is bounded~by 
\[
M=\sum_{g=0}^Gg=\frac{G(G+1)}2.
\]
A semigroup $S$ is represented by a quadruple $(c(S),g(S),m(S),\delta_S)$. 
By relations $ii)$ and $iii)$ of Proposition~\ref{P:Res}, we have $c\leq 2g(S)$ and $m\leq g(S)+1$. 
As $g(S)\leq G$ holds, the integers~$c$, $g$  and $m$ of the representation of $S$ require a memory space in $O(\log(G))$. 
The size of~$\delta_S=(d_S(0),...,d_S(3G))$ is exactly~$3G+1$. 
Each entry of $\delta_S$ is the $S$-decomposition number of an integer smaller than $3G$ and hence
requires $O(\log(G))$ bytes of memory space.
Therefore the space complexity of $\delta_S$ is in~$O(\log(G)\times G)$, which implies that the space complexity of the \textsc{Count} algorithm is
\[
O(\log(G)\times G\times M)= O(\log(G)\times G^3).\qedhere
\]
\end{proof}

@@

\item Vectorization (\MMX, \SSE instructions sets) and careful memory alignment; 
\item Shared memory multi-core computing using \CilkP for low level
  enumerating tree branching;
\item Partially derecursived algorithm using a stack;
\item Avoiding all dynamic allocation during the computation: everything is
  computed ``in place'';
\item Avoiding all unnecessary copy (the cost of the \textsc{Son} algorithm is
  roughly the same as copying);
\item Aggressive loop unrolling: the main loop is unrolled by hand using some
  kind of Duff's device;
\item Careful choice of data type (\verb|uint_fast8_t| for decomposition
  number, vs \verb|uint_fast64_t| for all indexes).
\end{itemize}

The source code of our algorithm is available in \cite{code}.
\subsection{Vectorization}

@@

{\small
\begin{alltt}
start = x >> 4;      // index of the block containing x
shift = x & 0xF;     // offset of x inside the block
...                  // some specific instructions to handle the first incomplete block.
for (long int i=start+1; i<NBLOCKS; i++) {
  block = load_unaligned_epi8(src + ((i-start)<<4) - shift);
  dst[i] -= ((block != zero) & one);
}
\end{alltt}
}
